% Canonical source for the mirror-protocol complexity analysis (commit
% e6471062, after the 16-byte Merkle-hash truncation in d4eeb8f).
% Supersedes the retired markdown original
% (results/mirror-complexity.md).
\documentclass[11pt]{article}

\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage[colorlinks=true, linkcolor=blue, urlcolor=blue]{hyperref}

% Monospace treatment for code identifiers and file:line references.
% Allow line breaks after path separators so long file:line references
% do not overflow the margin.
\newcommand{\code}[1]{\texttt{#1}}
\emergencystretch=3em
% The unbreakable tt file:line blobs leave a handful of lines loose; that
% is accepted layout here, so do not report underfull hboxes (overfull
% detection is unaffected).
\hbadness=10000

\theoremstyle{plain}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem*{corollary}{Corollary (separation depth)}

\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}

\title{The asymptotic complexity of the mirror reconciliation protocol}
\author{}
\date{\normalsize\itshape Canonical source; supersedes the retired markdown
original (\code{results/mirror-complexity.md}). Revised 2026-06-12 for the
two-width hash construction (16-byte Merkle, 32-byte content; commit
\code{d4eeb8f}).}

\begin{document}
\maketitle

A first-principles derivation of the per-side computation, bandwidth, and
latency of \code{rumors}' mirror protocol, written for a reader who knows
basic probability and big-O but has never seen this codebase. Every modeling
assumption is pinned to a code location; every approximation is named where
it is used. All file references are relative to the repository root (this
worktree); line numbers are as of commit \code{e6471062}.

This revision reflects the two-width hash construction (commit
\code{d4eeb8f}): every subtree hash the protocol ships and compares is a
\textbf{16-byte} truncated BLAKE3 (the Merkle width), while content
hashes --- the leaf paths --- remain full-width \textbf{32 bytes} (\S10.7
states the asymmetry argument). Every \emph{count} in this analysis ($W$,
$\Delta_k$, round trips) is width-independent and unchanged; the
per-record byte constants roughly halve.

\paragraph{Headline results} (derived in \S6, constants from the code):

\begin{center}
\begin{tabular}{@{}L{0.16\textwidth}L{0.46\textwidth}L{0.28\textwidth}@{}}
\toprule
Axis & Result (expected, uniform leaves) & Empirical fit it refines \\
\midrule
Per-side computation &
$\Theta\bigl(\min(256 \cdot D,\, N) \cdot (1 + \log_{256}(N/D))\bigr)$
child-hash comparisons; equivalently $\Theta(D \cdot (1 + \log(N/D)))$
disputed nodes visited, $\times 256$ sibling fan-out each &
$\sqrt{C \cdot D}$: a true upper bound (\S7), loose by up to
$\sqrt{C/D}/\ln(C/D)$ \\
\addlinespace
Bandwidth &
$\Theta(D)$ message bodies + the same
$\min(256 \cdot D,\, N) \cdot (1 + \log_{256}(N/D))$ hash entries
(${\sim}18$--$48$\,B each) + $\le 38$ frames of framing &
${\sim}D$: correct up to the slowly-varying log factor \\
\addlinespace
Latency &
$\tfrac{1}{2} \cdot \log_{256}(2 \cdot D \cdot N) + O(1)$ round trips, hard
ceiling $\approx 19$ RTT (38 frames), floor 1 RTT after the preamble for
identical versions &
$\log_{256}(C + D)$: same at $D \approx N$, ${\sim}2\times$ high for
$D \ll N$ (both tiny) \\
\bottomrule
\end{tabular}
\end{center}

where $N = C + D$. One meta-finding up front, developed in \S7: at every
realistic scale ($N \le 10^7$), the \emph{exact} level-by-level sum tracks
$\sqrt{C \cdot D}$ to within a factor of ${\sim}2$--$4$ across almost the
whole $(C, D)$ plane. The smooth $D \cdot \log(N/D)$ form is the asymptote;
the discrete base-256 staircase underneath it locally mimics a square root.
The empirical fit was not an accident, and wall-clock benchmarks at these
scales genuinely cannot refute it --- \S9 gives measurements that can.

\begin{quote}
\textbf{Source-of-data flag.} \code{benches/hash\_recon.rs} is \emph{not} a
reconciliation benchmark: it measures worst-case root-hash
\emph{reconstruction} (32 iterated 8\,KiB BLAKE3 wraps vs.\ 34-byte
commitments), motivating the per-node hash memo. Note it still models the
\emph{pre-truncation} 32-byte child records; the shipped tree's constants
are now ${\approx}4.3$\,KiB and 18\,B (\S10.1), so its absolute numbers
are ${\sim}2\times$ today's, the comparison's shape unaffected. The
reconciliation measurements live in \code{benches/gossip\_grid.rs} and
\code{benches/in\_memory.rs} over the divergence grid in
\code{benches/support/grid.rs}. Everything below assumes the
$\sqrt{C \cdot D}$ fit came from that grid.
\end{quote}

\section{The system, in one page}

Each replica holds a \textbf{sparse Merkle radix trie}: branching factor
256, fixed depth 32, one path byte per level (\code{src/tree.rs:7-26},
\code{src/tree/typed/height.rs:108-125}). A leaf's 32-byte path is
$\mathrm{blake3}\bigl(\mathrm{blake3}(\textit{version}) \,\Vert\,
\mathrm{blake3}(\textit{value})\bigr)$ --- the \emph{version is folded into
the path} (\code{src/tree/typed/path.rs:32-50}), so every insert, even of
identical content, lands at a fresh, computationally-uniform position. Two
hash widths are in play, deliberately: the path is a full-width 32-byte
\emph{content} hash, while every subtree hash the protocol ships and
compares is a 16-byte truncated \emph{Merkle} hash
(\code{MERKLE\_HASH\_LEN}, \code{src/tree/typed/hash.rs:11}; the asymmetry
argument is \S10.7).
Single-child runs are path-compressed into a prefix byte string on the node
(\code{src/tree/typed/untyped.rs:29-51}, invariant ``branches have $\ge 2$
children'' at \code{untyped.rs:82-84, 129-154}). Each node lazily memoizes
its subtree hash (\code{untyped.rs:261-292}) and its version ceiling/floor
(\code{untyped.rs:294-364}); the memos survive copy-on-write clones
(\code{untyped.rs:104-127}). Hashing is compression-invariant: a compressed
prefix byte wraps the child hash exactly as a materialized one-child branch
would (\code{src/tree/typed/hash.rs:101-126}, \code{untyped.rs:281-289}).
Leaves hash to a constant (\code{hash.rs:71-75, 92-98}) --- all
leaf-distinguishing content lives in the \emph{path}.

The \textbf{mirror protocol} reconciles two replicas' trees over a wire.
After a framed preamble (29 bytes on the wire: a 4-byte length prefix over
magic, protocol version, network id, and intent, exchanged concurrently by
both sides; \path{src/tree/traverse/mirror/remote/preamble.rs:5-15,66-72})
and a framed version handshake, equal versions terminate immediately
(\code{src/tree/traverse/mirror/local.rs:236-245, 274-285};
\code{remote.rs:269-275}). Otherwise both sides run a synchronized descent.
Each side keeps a \emph{zipper} of per-height levels
(\code{src/tree/typed/levels.rs}); each message a side sends descends its
own zipper by \textbf{two heights} (\code{local.rs:140-143};
\code{partition.rs:353-356}), and because the two sides alternate strictly,
the conversation's frontier advances \textbf{one depth per message, two per
round trip}.

Each steady-state message carries three channels, per the wire format in
\path{src/tree/traverse/mirror/message.rs:52-72} and the asymmetry matrix
at \code{local.rs:47-59}:

\begin{center}
\begin{tabular}{@{}L{0.14\textwidth}L{0.32\textwidth}L{0.44\textwidth}@{}}
\toprule
Channel & Content & Receiver's action \\
\midrule
\code{uncertain} &
$(\textit{prefix}, \textit{hash})$ of the sender's disputed-frontier
subtrees &
merge-join against own children: equal hash $\Rightarrow$ drop; differ
$\Rightarrow$ explode grandchildren (recurse); we-lack $\Rightarrow$
\code{request}; they-lack $\Rightarrow$ \code{provide} \\
\addlinespace
\code{requested} &
prefixes the sender lacks entirely &
answer next message by exploding that node's children into \code{providing}
(\code{partition.rs:91-142}) \\
\addlinespace
\code{providing} &
whole subtrees, full structure on the wire (\code{message.rs:52-60}) &
insert at the named prefix (\code{partition.rs:57-85}) \\
\bottomrule
\end{tabular}
\end{center}

All per-round set operations are single linear merges over sorted flat
vectors (\code{partition.rs:}\allowbreak\code{100-115, 197-203};
\code{src/tree/typed/}\allowbreak\code{levels/level.rs:1-10}), so per-round work
is linear in the
frontier and message sizes --- no hidden log factors beyond the \code{imbl}
\code{OrdMap} child maps (an $O(\log 256)$ constant per child operation,
absorbed below).

\textbf{Termination is in-band}: a side is done when its outgoing
\code{requested} and \code{uncertain} are both empty
(\code{partition.rs:393-410}; \code{remote.rs:38-46, 354-364, 388-399}); a
sender that just emitted an empty-empty message does not even await a reply
(\code{remote.rs:388-399}). The schedule has a hard ceiling ---
\code{mirror\_connected} (\code{src/tree/traverse/mirror.rs:164-185}) is
\code{initiator} + \code{responder} + \code{open\_initiator} + a \code{seq!}
loop of \textbf{14 exchange pairs} + \code{close\_initiator} +
\code{complete\_responder} + \code{complete\_initiator}. Counting frames:
2 (concurrent) preamble frames + 2 handshake + \code{Initiate} (depth 0) +
\code{Opening} (depth 1) +
\code{Exchange} (depth 2) + 28 exchanges (depths 3--30) + one more responder
exchange (depth 31) + \code{Closing} + \code{Complete} = \textbf{38 frames
maximum} --- the 36-frame schedule plus the framed preamble pair ---
covering all 32 byte-levels (the chain lengths are pinned at the
type level in \code{src/tree/traverse/mirror/protocol.rs:497-508}).

\textbf{Deletions} leave no tombstones. Every subtree shipped through
\code{providing} is first filtered by \code{Unknown::unknown}
(\code{src/tree/traverse/unknown.rs}): a subtree whose version
\emph{ceiling} is $\le$ the counterparty's version vector is something they
have already seen --- its absence on their side means they deleted it --- so
it is dropped in $O(1)$ (one memoized-ceiling comparison,
\code{unknown.rs:39-47}), \emph{and the local copy is dropped too} (the
Left arm keeps a node only if it survives the filter,
\code{partition.rs:232-239}). A floor check keeps wholly-unknown subtrees
intact without rebuilding them (\code{unknown.rs:49-66}).

\section{Model and assumptions}

\begin{itemize}
\item \textbf{(A1) Uniform leaf placement.} Leaf paths are BLAKE3 outputs
  over $(\textit{version}, \textit{value})$ (\code{path.rs:32-50}), and
  versions are unique per action (\code{src/tree.rs:311-323, 341-348}). In the
  random-oracle model the $N$ leaf paths are i.i.d.\ uniform 32-byte
  strings, \emph{regardless of insertion order, batching, or payload
  skew} --- an adversary would need structured BLAKE3 preimages to cluster
  them. This is the load-bearing probabilistic assumption; see \S10.4.
\item \textbf{(A2) Parameters.} Branching $b = 256$, depth 32
  (\code{tree.rs:9}, \code{height.rs:108-125}). Two hash widths: the
  Merkle (comparison) hash is $h = 16$ bytes, a prefix truncation of
  BLAKE3 (\code{MERKLE\_HASH\_LEN}, \code{hash.rs:11}); the content hash
  --- the leaf path --- stays 32 bytes (\code{ContentHash},
  \code{hash.rs:153-165}; \code{path.rs:32-50}).
\item \textbf{(A3) Warm memos.} Subtree hashes and version bounds are
  memoized reads (\code{untyped.rs:261-292}); we count a hash
  \emph{comparison} as $O(1)$. First-touch costs after local mutation are a
  separate, additive term (\S10.1) --- the benches warm them in untimed
  setup (\code{benches/gossip\_grid.rs:119-160}, \code{src/tree.rs:241-256}).
\item \textbf{(A4) Dispute drives descent.} Only the
  \emph{Both-and-hashes-differ} cell of the asymmetry matrix recurses
  (\code{partition.rs:243-254}). The Left cell ships a whole subtree in the
  same message; the Right cell is requested and answered whole one message
  later (\code{partition.rs:91-142}). So one-sided differences cost $O(1)$
  additional messages once they separate from two-sided content.
\item \textbf{(A5) Two heights per send, one depth per message.} Verified
  at \code{partition.rs:}\allowbreak\code{353-356} and the alternation in
  \code{mirror.rs:}\allowbreak\code{164-185}.
\item \textbf{(A6) No level skipping.} Path compression does \textbf{not}
  shorten the descent: exploding a compressed node pops exactly one prefix
  byte and yields a singleton child map (\code{untyped.rs:160-170}).
  Compression saves hashing and memory, never rounds.
\item \textbf{(A7) No hash collisions.} A false \code{uncertain} match
  (probability $2^{-128}$ \emph{per comparison} at the 16-byte Merkle
  width --- pairwise, never birthday-amplified, since the hash at prefix
  $P$ only ever meets the counterparty's hash at the same $P$) would
  silently prune one divergent subtree until the next mutation beneath
  that prefix forces a re-compare. Cumulatively ${\approx}2^{-76}$ over a
  paranoid fleet lifetime (derived in \code{hash.rs:24-58}); ignored.
\item \textbf{(A8) Wire sizes.} An \code{uncertain} entry at depth $k$ is
  $k$ raw prefix bytes + 16 hash bytes (\code{prefix.rs:127-142},
  \code{message.rs:7-21, 63-65}); a \code{providing} node is its full borsh
  structure: per leaf, a version, the payload, and $\le 32$
  compressed-prefix bytes --- paths keep the full content width
  (\code{message.rs:29-50}, \code{untyped.rs:383-443}). Frames carry a
  4-byte length prefix (\code{remote.rs:25-36}).
\end{itemize}

\section{Notation}

\begin{center}
\begin{tabular}{@{}lL{0.72\textwidth}@{}}
\toprule
Symbol & Meaning \\
\midrule
$b$ & branching factor, 256 \\
$C$ & live messages present on \textbf{both} sides \\
$D$ & messages present on \textbf{exactly one} side, summed over both sides
  ($D = D_L + D_R$); redacted-on-one-side leaves count (\S10.3) \\
$N$ & $C + D$, the size of the union tree \\
$k$ & depth = prefix length in bytes, $0 \le k \le 32$; height $H = 32 - k$ \\
$\Delta_k$ & number of \textbf{disputed} depth-$k$ prefixes: prefixes
  containing $\ge 1$ differing leaf \emph{and} $\ge 1$ leaf on the
  counterparty's side \\
$f_k$ & expected occupied-children count of a depth-$k$ node,
  $\approx \min(b, \max(1, N/b^k))$ \\
$j_D$, $j_N$ & $\log_b D$, $\log_b N$ (depths where $b^k$ crosses $D$, $N$) \\
$k^*$ & separation depth: deepest disputed prefix \\
$W$ & per-side work / total \code{uncertain} entries:
  $\sum_k \Delta_k \cdot f_k$ \\
$\bar{s}$, $s_v$ & mean payload size; serialized \code{Version} size (not
  $O(1)$, \S10.5) \\
\bottomrule
\end{tabular}
\end{center}

Logs are base $b = 256$ unless written $\ln$. One decimal decade is
$\log_{256} 10 \approx 0.415$ levels; one level spans 2.4 decades. Keep that
coarse quantization in mind --- it is the source of the staircase in \S7.

\section{Warm-up: \texorpdfstring{$D = 1$, $N = 10^6$}{D = 1, N = 10\textasciicircum 6}}

One side holds a single extra leaf; the trees otherwise agree on $10^6$
messages. Trace the protocol:

\begin{enumerate}
\item \textbf{Handshake} (2 frames): versions differ, descend.
\item \textbf{\code{Initiate}} (depth 0): one root hash. Differs.
\item \textbf{\code{Opening}} (depth 1): the responder \emph{unconditionally}
  explodes its root and lists all its children (\code{local.rs:338-357}) ---
  at $N = 10^6$, all 256 slots are occupied ($\approx 3{,}906$ leaves each):
  \textbf{256 entries}.
\item The initiator merge-joins: 255 child hashes match and drop; the one
  slot containing the extra leaf differs (Both-case), so its 256
  grandchildren are exploded and listed: \textbf{256 entries} at depth 2.
\item The responder finds one depth-2 hash differing; that node has
  $\approx 10^6/256^2 \approx 15$ occupied children:
  \textbf{$\approx$ 15 entries} at depth 3.
\item At depth 3 the extra leaf sits alone in its slot with probability
  $\approx 1 - N/256^3 \approx 0.94$: a \textbf{Left case}. The whole leaf
  (a path-compressed spine node carrying its remaining $\approx 29$ path
  bytes, version, and payload) ships in \code{providing}; nothing is
  exploded; the outgoing \code{uncertain} and \code{requested} are empty
  $\Rightarrow$ \textbf{Done}, in-band.
\end{enumerate}

Totals: $\approx 256 + 256 + 15 \approx 527$ hash entries on the wire and
compared (${\sim}9$\,KB at $k + 16$ bytes per entry), 7--8 frames
$\approx$ 4 round trips including the preamble, one message body. Three things to notice, which the general
analysis formalizes:

\begin{itemize}
\item The \emph{visited disputed nodes} number only $\approx 3$ (one per
  level until separation) --- $D \cdot (1 + \log_b(N/D))$ with $D = 1$.
\item But each visited node costs its \textbf{full sibling fan} ($\le 256$
  entries): the per-node constant is $f_k \approx \min(b, N/b^k)$. This
  factor-256 is the price of the radix-256 tradeoff: few rounds, fat rounds.
\item Separation happened at depth
  $\approx \log_b(D \cdot N) = \log_b 10^6 \approx 2.5$ --- not
  $\log_b N$ levels of \emph{messages saved}, because termination is the
  in-band emptiness predicate, not bottoming out at depth 32.
\end{itemize}

\section{Two lemmas}

\begin{lemma}[prefix occupancy / sharing]
Throw $m$ uniform leaves into the $b^k$ bins at depth $k$. The expected
number of occupied bins is $b^k \cdot (1 - (1 - b^{-k})^m)$, which is
$\Theta(\min(b^k, m))$ --- at least $(1 - 1/e) \cdot \min(b^k, m)$, at most
$\min(b^k, m)$.
\end{lemma}

\begin{proof}[Proof sketch]
Linearity of expectation over bins; $(1 - 1/x)^x \le 1/e$. Applied to the
$D$ differing leaves, this bounds how many distinct ancestors they have at
each depth: their root-to-leaf paths share prefixes exactly while
$b^k < D$, and are essentially disjoint below. Summing, the union of the
$D$ differing root-to-separation paths has
$\sum_k \min(b^k, D) = \Theta(D \cdot (1 + \log_b(N/D)))$ distinct nodes
(the geometric head contributes $\Theta(D)$; the plateau contributes $D$
per level from $j_D$ to the dispute cutoff of Lemma~2).
\end{proof}

\begin{lemma}[dispute persistence]
A depth-$k$ prefix is disputed --- i.e.\ recursed into rather than
resolved --- iff it contains a differing leaf \emph{and} the counterparty
owns at least one leaf under it (A4: otherwise it is a Left/Right case and
resolves in $O(1)$ messages). Under A1,
\begin{equation*}
\mathbb{E}[\Delta_k] \;\le\; \min\bigl( b^k,\; D,\; 2 \cdot D \cdot N \cdot b^{-k} \bigr)
\end{equation*}
and $\Pr(\text{any dispute survives depth } k) \le 2 \cdot D \cdot N \cdot b^{-k}$.
\end{lemma}

\begin{proof}[Proof sketch]
The first two bounds are Lemma~1 applied to the differing leaves. For the
third: a dispute at depth $k$ requires some differing leaf $d$ and some
leaf $x$ on the side opposite $d$'s holder to share a $k$-byte prefix. Each
such (ordered) pair shares it with probability $b^{-k}$ (uniform,
independent paths), and there are at most
$D_L \cdot (C + D_R) + D_R \cdot (C + D_L) \le 2 \cdot D \cdot N$ pairs.
Union bound / linearity.
\end{proof}

\begin{corollary}
$k^* \le \log_b(2 \cdot D \cdot N) + t$ with probability $\ge 1 - b^{-t}$;
$\mathbb{E}[k^*] \le \log_b(2DN) + \frac{b}{b-1} \cdot \frac{1}{\ln b}
\approx \log_b(2DN) + 1$. The fixed 32-level ceiling is unreachable
probabilistically: hitting it would need $D \cdot N \approx 256^{30}$. (It
exists because paths are 32 bytes; even a full-depth descent is
\emph{correct} --- equal full paths imply equal
$(\textit{version}, \textit{value})$ by construction, \code{tree.rs:77-82}.)
\end{corollary}

\begin{lemma}[fan-out]
A depth-$k$ node in the union tree of $N$ uniform leaves has expected
occupied-children count $f_k = \Theta(\min(b, \max(1, N/b^k)))$, and the
\emph{total} children over any set of $\Delta$ depth-$k$ nodes is
$\le \Delta \cdot b$ always and $\Theta(\Delta \cdot f_k)$ in expectation
(children counts across disjoint subtrees are negatively associated; we
only ever need linearity). \emph{Caution (Jensen):}
$\mathbb{E}[\min(X, Y)] \le \min(\mathbb{E}X, \mathbb{E}Y)$, so products of
the $\min$ envelopes used below are upper envelopes of expectations; the
matching lower bounds hold within the $(1 - 1/e)$ constants of Lemma~1.
\end{lemma}

\section{The level-by-level sum}

Per round at depth $k$, the side processing the counterparty's
\code{uncertain} does (all single-pass, sorted merges ---
\code{partition.rs:159-304}):

\begin{itemize}
\item drain its frontier once: $O(|\textit{frontier}_k|)$;
\item for each disputed parent, merge-join its own occupied children
  against the listed hashes: $O(\text{own fan} + \text{listed entries})$,
  each comparison a memoized-hash \code{memcmp} (A3);
\item explode each hash-differing child's grandchildren into the next
  level and emit them as the next \code{uncertain}
  (\code{partition.rs:244-254, 365-373}).
\end{itemize}

So the entries \emph{listed at depth $k$} are exactly the occupied children
of the disputed depth-$(k-1)$ parents, and each side's per-session work,
the total \code{uncertain} traffic, and the zipper/\code{collapse} churn
(\code{levels.rs:154-203}, linear in total level entries) are all $\Theta$
of the same master sum:
\begin{equation*}
W \;=\; \sum_{k=0}^{32} \Delta_k \cdot f_k
\;\approx\; \sum_k \min\bigl(b^k,\, D,\, 2DN \cdot b^{-k}\bigr)
            \cdot \min\bigl(b,\, \max(1,\, N/b^k)\bigr)
\end{equation*}

Evaluate by regimes (write $j_D = \log_b D$, $j_N = \log_b N$; assume first
$D \le N/b^2$ so all three regimes are non-empty):

\begin{center}
\begin{tabular}{@{}L{0.30\textwidth}lL{0.13\textwidth}L{0.26\textwidth}@{}}
\toprule
Regime & Depths & Term & Contribution \\
\midrule
\textbf{Head} (paths shared) & $k \le j_D$ & $b^k \cdot b$ &
  geometric, $\approx b \cdot D \cdot (1 + 1/b + \dots) \approx b \cdot D$ \\
\addlinespace
\textbf{Plateau} (paths distinct, fans saturated) & $j_D < k \le j_N - 1$ &
  $D \cdot b$ & $b \cdot D \cdot (\log_b(N/D) - 1)$ \\
\addlinespace
\textbf{Tail} (fans thin out, disputes die) & $k > j_N - 1$ &
  $D \cdot N \cdot b^{-k} \cdot O(1)$ &
  geometric, $\approx \Theta(D \cdot b)$ at the shoulder, then $\div 256$
  per level \\
\bottomrule
\end{tabular}
\end{center}

\begin{theorem}[per-side computation]
Under A1--A6, the expected per-side work (hash comparisons + level-entry
handling, each $O(1)$) and the expected total \code{uncertain} entry count
are
\begin{equation*}
W \;=\; \Theta\bigl( \min(b \cdot D,\, N) \cdot (1 + \log_b(N/D)) \bigr)
\end{equation*}
\begin{itemize}
\item for $D \le N/b$: $W = \Theta(b \cdot D \cdot (1 + \log_b(N/D)))$ ---
  the prompt-level form $\Theta(D \cdot (1 + \log(N/D)))$ counts
  \emph{disputed nodes visited}; the protocol pays a $\times b$ sibling fan
  on each (Lemma~3, \S4);
\item for $N/b < D \le N$ (heavily diverged / disjoint): the plateau
  vanishes and the peak term is $\Theta(N)$;
  $W = \Theta(N) = \Theta(D) \cdot O(b)$ --- linear;
\item additive $\Theta(D \cdot (1 + s_v))$ for absorbing/filtering the
  provided subtrees and the final collapse, and $\Theta(1)$ per round of
  fixed pipeline cost. \qed
\end{itemize}
\end{theorem}

Cross-checks: for $D = 1$, $N = 10^6$ the log factor sharpens from the
envelope's $\log_b(N/D)$ to $\log_b(N/b)$, because the per-level fan $f_k$
saturates at $b$ only down to depth $j_N - 1$ and decays geometrically
below --- so the sum is $b$ per level over the $j_N = 1 + \log_b(N/b)$
saturated levels, $W \approx 256 \cdot (1 + \log_b(N/b)) \approx 640$, vs.\
the \S4 hand count of 527 \checkmark{} (the residue is the envelope billing
the thinned final fraction of a level at the full $b$ where the tree
delivers ${\approx}15$). Billing all $1 + \log_b(N/D) \approx 3.5$ levels of
descent at full fan, as the $\Theta$-form read literally would, gives
${\approx}894$ --- still a correct upper envelope, but the descent below
depth $j_N - 1$ is exactly where the fans thin. Disjoint trees ($C = 0$,
$N = D$) give $\Theta(N)$ \checkmark{} (matches the one-shot character of
bootstrap, \S10.2).

\begin{theorem}[bandwidth]
Total bytes on the wire, both directions (each depth's \code{uncertain} is
listed by exactly one side --- the alternation splits levels between them):
\begin{align*}
\textit{bytes} \;=\;& W \cdot (h + \bar{k})
  && \text{[\code{uncertain}: 16-byte hash + $k$-byte prefix, $\bar{k} \le j_N + 1$]} \\
  &+ \Theta(D) \cdot (\bar{s} + s_v + {\le}32)
  && \text{[\code{providing}: payloads, versions, compressed spines]} \\
  &+ O(D) \cdot \bar{k}
  && \text{[\code{requested}: one prefix per separated they-only subtree]} \\
  &+ {\le}\, 36 \cdot 4 + 2 \cdot 29 + 2 \cdot s_v
  && \text{[framing, preamble, handshake]}
\end{align*}
In $\Theta$-terms:
$\Theta(D \cdot (\bar{s} + s_v))
 + \Theta(\min(b \cdot D, N) \cdot (1 + \log_b(N/D)) \cdot h)$.
The empirical ``${\sim}D$'' is right in its $D$-scaling; the per-$D$
constant is the interesting part: at $N/D = 10^2 \dots 10^4$ the smooth
envelope prices the hash channel at ${\sim}5$--$20$\,KB \emph{per
differing message} (e.g.\ \S4: 9\,KB to move one message). That envelope
is tight only for $D \lesssim b$; \S8 prices the channel exactly and shows
the staircase discounts it to ${\sim}0.2$--$3$\,KB once $D \gg b$. For
small payloads, \textbf{hash traffic dominates bandwidth}; bodies dominate
only when $\bar{s} \gtrsim 20 \cdot W/D$ bytes (the crossover is solved in
\S8). \qed
\end{theorem}

\begin{theorem}[latency]
Messages strictly alternate and the conversation advances one depth per
message (A5). Disputes empty at depth $k^*$ (Lemma~2 corollary), plus
$O(1)$ messages to flush the final
$\code{requested} \to \code{providing}$ exchange; the in-band predicate
then stops the session (A6 --- no draining of the remaining schedule):
\begin{equation*}
\mathrm{RTT} \;=\; \tfrac{1}{2} \cdot k^* + O(1)
\;=\; \tfrac{1}{2} \cdot \log_{256}(2 \cdot D \cdot N) + c,
\qquad c \approx 3\text{--}4
\end{equation*}
(preamble $\approx 1$, handshake 1, \code{Initiate}/\code{Opening} 1,
closing tail ${\sim}1$), hard-ceilinged at 38 frames $\approx$ 19 RTT
(the concurrent preamble pair included),
floored at ${\sim}2$ RTT total for equal versions. Path compression does
not reduce this (A6). Numerically the \emph{descent} portion is tiny at any
scale --- 1.8 levels $\approx$ 1 RTT for $(N, D) = (10^6, 1)$ up to 6.1
levels $\approx$ 3 RTT for $(10^7, 10^7)$; the $O(1)$ dominates. The
empirical $\log_{256}(C + D)$ agrees at $D \approx N$ (where
$\tfrac{1}{2} \log DN = \log N$) and is $\approx 2\times$ the true descent
depth for $D \ll N$, a difference of $< 2$ RTT at any realistic scale ---
unfalsifiable from timing, trivially checkable by counting frames (\S9). \qed
\end{theorem}

\section{Against the \texorpdfstring{$\sqrt{C \cdot D}$}{sqrt(C·D)} fit}

\begin{proposition}[the fit is a true upper bound]
For $1 \le D \le C$:
\begin{equation*}
D \cdot \ln(C/D) \;\le\; \frac{2}{e} \cdot \sqrt{C \cdot D},
\qquad \text{with equality iff } D = C/e^2.
\end{equation*}
\end{proposition}

\begin{proof}
Set $x = D/C \in (0, 1]$ and $g(x) = \sqrt{x} \cdot \ln(1/x)$. On $(0, 1]$,
$g'(x) = (\ln(1/x) - 2)/(2\sqrt{x})$ vanishes only at $x = e^{-2}$, a
maximum, with $g(e^{-2}) = 2/e$. Since $\sqrt{CD} = C\sqrt{x}$,
\begin{equation*}
D \cdot \ln(C/D)
\;=\; C \cdot x \cdot \ln(1/x)
\;=\; (C\sqrt{x}) \cdot \bigl(\sqrt{x} \cdot \ln(1/x)\bigr)
\;=\; \sqrt{CD} \cdot g(x)
\;\le\; \frac{2}{e} \cdot \sqrt{CD},
\end{equation*}
with equality iff $x = e^{-2}$, i.e.\ $D = C/e^2$.
\end{proof}

In base 256:
$D \cdot \log_{256}(C/D) \le (2/(e \cdot \ln 256)) \cdot \sqrt{CD}
\approx 0.133 \cdot \sqrt{CD}$. With Theorem~1's constants
($b/\ln b \approx 46.2$), for $D \le C$:
\begin{equation*}
W \;\le\; b \cdot D \cdot (2 + \ln 2)
        + \frac{2}{e} \cdot \frac{b}{\ln b} \cdot \sqrt{CD}
\;\approx\; 690 \cdot D + 34 \cdot \sqrt{C \cdot D}
\;=\; O(\sqrt{C \cdot D}),
\end{equation*}
since $D \le \sqrt{CD}$ when $D \le C$. So the empirical
$\sqrt{C \cdot D}$ is a \emph{correct} upper bound on per-side
computation --- but loose: the ratio
$\sqrt{CD}/(D \cdot \log(C/D)) = \sqrt{C/D}/\log(C/D)$ grows without bound
as $D/C \to 0$, and the bound is tight (constant-factor) only in the band
$D/C \in [{\sim}10^{-2}, {\sim}\tfrac{1}{2}]$ where $\sqrt{x} \cdot \ln(1/x)$
stays within $1.5\times$ of its max.

\paragraph{Why the fit nevertheless matched the data: the staircase.}
The smooth form $b \cdot D \cdot (1 + \log_b(N/D))$ is the envelope of a
\emph{staircase}: each level of the sum spans a factor of $256 \approx 2.4$
decades. Inside one span the dominant term is $t_k = D \cdot N/b^k$ ---
\textbf{linear in $N$ at fixed $D$} --- and the term above it,
$\min(b, D) \cdot b^k$-shaped, is \emph{flat} in $D$ once $D > b$. The
exact sum's local log-log slopes therefore wobble around the sqrt's 0.5
instead of sitting at the asymptotic 0 (in $N$) and 1 (in $D$):

Exact-sum values of $W$ (children examined $\approx$ \code{uncertain}
entries; computed from the master sum, expectations, constants $\pm 2\times$
from occupancy corrections), against $\sqrt{C \cdot D}$:

\begin{center}
\begin{tabular}{@{}rrrrrrr@{}}
\toprule
$N$ & $D$ & $W$ & $\sqrt{C \cdot D}$ & $W/\sqrt{CD}$ & $k^*$ & RTT (descent${}+c$) \\
\midrule
$10^4$ & $1$      & $295$              & $100$              & $3.0$            & $1.8$ & ${\sim}3.9$ \\
$10^4$ & $10^2$   & $4.2\,\mathrm{k}$  & $1.0\,\mathrm{k}$  & $4.2$            & $2.6$ & ${\sim}4.3$ \\
$10^4$ & $10^4$   & $26\,\mathrm{k}$   & $10\,\mathrm{k}$   & $2.6$            & $3.6$ & ${\sim}4.8$ \\
$10^5$ & $10^2$   & $26\,\mathrm{k}$   & $3.2\,\mathrm{k}$  & $\mathbf{8.2}$   & $3.0$ & ${\sim}4.5$ \\
$10^5$ & $10^3$   & $67\,\mathrm{k}$   & $10\,\mathrm{k}$   & $6.8$            & $3.5$ & ${\sim}4.7$ \\
$10^6$ & $1$      & $0.53\,\mathrm{k}$ & $1.0\,\mathrm{k}$  & $\mathbf{0.53}$  & $2.5$ & ${\sim}4.2$ \\
$10^6$ & $10^2$   & $27\,\mathrm{k}$   & $10\,\mathrm{k}$   & $2.7$            & $3.5$ & ${\sim}4.7$ \\
$10^6$ & $10^3$   & $81\,\mathrm{k}$   & $32\,\mathrm{k}$   & $2.6$            & $3.9$ & ${\sim}4.9$ \\
$10^6$ & $10^4$   & $220\,\mathrm{k}$  & $100\,\mathrm{k}$  & $2.2$            & $4.3$ & ${\sim}5.1$ \\
$10^6$ & $10^5$   & $1.08\,\mathrm{M}$ & $316\,\mathrm{k}$  & $3.6$            & $4.7$ & ${\sim}5.3$ \\
$10^7$ & $1$      & $0.67\,\mathrm{k}$ & $3.2\,\mathrm{k}$  & $\mathbf{0.21}$  & $2.9$ & ${\sim}4.4$ \\
$10^7$ & $10^4$   & $1.6\,\mathrm{M}$  & $316\,\mathrm{k}$  & $5.1$            & $4.7$ & ${\sim}5.3$ \\
$10^7$ & $10^5$   & $10.2\,\mathrm{M}$ & $1.0\,\mathrm{M}$  & $\mathbf{10.2}$  & $5.1$ & ${\sim}5.6$ \\
$10^7$ & $10^7$   & $29\,\mathrm{M}$   & $10\,\mathrm{M}$   & $2.9$            & $6.1$ & ${\sim}6.0$ \\
\bottomrule
\end{tabular}
\end{center}

Read the ratio column: over five orders of magnitude in each axis it stays
within $[0.2, 10]$ --- a one-constant sqrt fit (say anchored at the central
cells, $a \approx 2.6$) leaves residuals of only $0.1\times$--$4\times$,
and the worst residuals sit at the extreme corners ($D = 1$ at huge $N$;
$D \approx N/b$, where the plateau has just vanished and the peak term
crosses from $b \cdot D$ to $N$) where wall time is dominated by fixed
per-session costs (preamble, handshake, ${\sim}8$ frames, thread/pipe
synchronization in the bench harness) or by the $\Theta(D)$ body-handling
term. Local log-log slopes of the exact sum, vs.\ sqrt's uniform 0.5:

\begin{itemize}
\item in $D$ at fixed $N = 10^6$: 0.75, 0.96, \textbf{0.47, 0.43}, 0.69
  per decade
  ($D\colon 1 \to 10 \to 10^2 \to 10^3 \to 10^4 \to 10^5$) --- \emph{in the
  densest benchmark range the true slope is $\approx 0.45$,
  indistinguishable from 0.5};
\item in $N$ at fixed $D = 10^3$: 0.80, \textbf{0.08}, 0.43 per decade
  ($N\colon 10^4 \to 10^5 \to 10^6 \to 10^7$) --- the $10^5 \to 10^6$
  decade is the smoking gun: the true cost is nearly flat there (it sits
  between staircase steps: $b^2 = 65{,}536$, $b^3 = 16.8\,\mathrm{M}$),
  while sqrt demands $\times 3.16$.
\end{itemize}

\paragraph{Conclusion.} $\sqrt{C \cdot D}$ and
$\Theta(\min(bD, N)(1 + \log_b(N/D)))$ are empirically indistinguishable
from wall time on diagonal or $D$-sweeps at grid scales; they separate
cleanly on \textbf{fixed-$D$, $N$-sweeps across the $10^5 \to 10^6$
decade}, and on any \emph{count}-based (rather than time-based)
measurement.

\section{Expected bandwidth in closed form}

Theorem~2 gave the bandwidth's shape. This section prices it: an explicit
expectation $E[B](C, D, M)$ --- total bytes on the wire, both directions,
to reconcile two replicas sharing $C$ live messages and differing in $D$
leaves, with mean serialized message body $M$. (A \code{Message<T>}
serializes byte-identically to its inner $T$, \code{message.rs:14-16}, so
$M$ is \S3's $\bar{s}$ as measured on the wire.) Split
$D = D_{\mathrm{live}} + D_{\mathrm{red}}$ into leaves the counterparty
must receive and leaves redacted on one side (\S10.3); $N = C + D$ as
always. Assumptions: A1 (uniform content addressing) and A8 (wire sizes);
the $D = D_L + D_R$ split between the sides turns out not to matter, shown
at the end.

\paragraph{\code{providing}: $33 + s_v + M$ bytes per live differing leaf.}
By A4 each differing leaf crosses the wire exactly once: a Left case ships
in the same message, a Right case is requested and answered whole one
message later (\code{partition.rs:91-142}), and disputed prefixes never
ship content, only hashes. A leaf provided alone at pairing depth $d$
costs, per the node encoding (\code{untyped.rs:383-443}; channel shapes at
\code{message.rs:52-66}): its pairing \code{Prefix}, $d$ raw bytes (the
typed height pins the byte count, so there is no length prefix,
\code{prefix.rs:127-142}); the node's \code{prefix\_len} byte plus its
$32 - d$ compressed path bytes (the spine from depth $d$ down to the
leaf); its \code{Version}, $s_v$ bytes on average; and its body, $M$
bytes. The pairing prefix and the compressed spine \emph{partition} the
leaf's 32 path bytes, so the path overhead is $d + 1 + (32 - d) = 33$
bytes --- independent of where the leaf separated. When several differing
leaves ship under one provided prefix, the shared spine bytes are paid
once and branch bookkeeping appears instead (one \code{prefix\_len} and
one \code{count\_minus\_two} byte per node, one radix byte per child): the
per-leaf path overhead $\ell$ then sits in $[2, 35]$, and sharing only
helps; the singleton value $\ell = 33$ is the common case, since uniform
paths rarely co-separate at $D \ll N$. Redacted leaves ship \emph{nothing}
here: both the Left arm and \code{answer\_requested} filter through
\code{Unknown::unknown} before anything is serialized
(\code{partition.rs:91-142, 232-239}; \code{unknown.rs:39-47}), so a
redacted leaf is billed for descent --- it counts in $D$, hence in
$W$ --- and at most one \code{requested} record, but zero body bytes.
In expectation:
\begin{equation*}
E[\code{providing}] \;=\; D_{\mathrm{live}} \cdot (\ell + s_v + M),
\qquad \ell \approx 33.
\end{equation*}

\paragraph{\code{uncertain}: $W$ records of $h + \bar{k}$ bytes.}
\S6 established that the records listed at depth $k$ are exactly the
occupied children of the disputed depth-$(k{-}1)$ parents: the wire
carries each disputed parent's \emph{full occupied sibling fan} --- the
Both arm explodes every grandchild, not only the differing ones
(\code{partition.rs:243-254}) --- and each depth is listed by exactly one
side (the alternation; Theorem~2). The session-total record count, both
directions summed, is therefore Theorem~1's $W$ itself:
\begin{equation*}
E[\#\text{records}] \;=\; W
\;=\; \Theta\bigl(\min(b \cdot D,\, N) \cdot (1 + \log_b(N/D))\bigr).
\end{equation*}
A record at depth $k$ is $k$ raw prefix bytes plus the $h = 16$-byte hash
(A8); $W$'s mass sits within a level of the saturation depth $j_N$ (the
head and shoulder of \S6's regime table), so the mean record size is
$\bar{r} = h + \bar{k}$ with $\bar{k} \le j_N + 1$: $\bar{r} \approx
18$--$22$ bytes for $N \le 10^7$. The numbers below use $\bar{r} = 20$.

\paragraph{\code{requested}: lower-order.} One bare prefix per separated
they-only subtree: $R$ records, $R \le D$ (each names a subtree holding at
least one differing leaf, and nothing is requested twice), each
$\le k^* + 1 \le 33$ bytes, no hash. So $\le 33 \cdot D$ bytes total ---
under the hash channel by a factor ${\sim}\bar{r} \cdot b/33 \approx 155$
in the dispute regime $D \le N/b$, and never above the bodies' fixed
per-leaf overhead. Lower-order everywhere.

\paragraph{Framing and handshake: $O(1)$.} Two 29-byte framed preambles
(4-byte header + 25-byte payload, \code{remote/preamble.rs:66-72}); two
framed \code{Handshake}s carrying one
\code{Version} each, $4 + s_v$ bytes framed (\code{message.rs:129-146});
at most 36 further frames per session (\S1) at 4 length bytes each
(\code{remote.rs:25-36}); and a \code{u32} length per channel vector, at
most three per frame (\code{message.rs:4-5}). Total fixed cost
\begin{equation*}
F \;\le\; 2 \cdot 29 + 2 \cdot (4 + s_v) + 36 \cdot 4 + 36 \cdot 12
\;\approx\; 640 + 2 \cdot s_v \text{ bytes},
\end{equation*}
independent of $C$, $D$, and $M$.

\paragraph{The expectation.} Summing the four contributions:
\begin{equation*}
E[B] \;=\;
\underbrace{D_{\mathrm{live}} \cdot (\ell + s_v + M)}_{\text{bodies}}
\;+\; \underbrace{W \cdot (h + \bar{k})}_{\text{hashes}}
\;+\; \underbrace{R \cdot \bar{k}_q}_{\text{requests}}
\;+\; F,
\end{equation*}
with $\ell \approx 33$, $h = 32$, $\bar{k} \le j_N + 1$, $R \le D$,
$\bar{k}_q \le 33$, $F = O(s_v)$. Dropping the lower-order terms and
absorbing the constants:
\begin{equation*}
E[B] \;=\; \Theta\Bigl( D_{\mathrm{live}} \cdot (M + s_v)
\;+\; \bar{r} \cdot \min(b \cdot D,\, N) \cdot
\bigl(1 + \log_b(N/D)\bigr) \Bigr),
\qquad \bar{r} = h + \bar{k} \approx 20.
\end{equation*}
Two caveats, both already on the record. The hash term is the smooth
envelope of \S7's staircase, so realized values mid-step run several times
below it. And when one side is empty (\S10.2) the dispute machinery never
engages: the hash term collapses to the \code{Opening}'s $O(b)$ records
and $E[B] \approx D \cdot (\ell + s_v + M) + F$ --- the formula's hash
envelope is then an overstatement, though the bodies dominate it there
regardless.

\paragraph{Crossover: when do bodies overtake hashes?} Set the two leading
terms equal and solve for $M$:
\begin{equation*}
M^* \;=\; \bar{r} \cdot \frac{W}{D_{\mathrm{live}}} \;-\; (\ell + s_v).
\end{equation*}
On the smooth envelope with $D \le N/b$ and $D_{\mathrm{red}} = 0$,
$W/D = b \cdot (1 + \log_b(N/D))$, so
\begin{equation*}
M^* \;\approx\; 20 \cdot 256 \cdot \bigl(1 + \log_{256}(N/D)\bigr)
\;\approx\; 9,\ 12,\ 14\ \text{KB at } N/D = 10^2,\ 10^3,\ 10^4
\end{equation*}
--- the source of Theorem~2's ``${\sim}5$--$20$\,KB per differing
message'' figure (the envelope, $\pm$ Lemma~1's occupancy constants). The
realized figures, taking $W$ from \S7's exact-sum table, agree with the
envelope only where it is tight, $D \lesssim b$: there
$M^* \approx 5$--$13$\,KB (the $D = 1$ rows and the $D = 10^2$ cells).
For $D \gg b$ the head levels' fans are shared among all $D$ differing
leaves and the shoulder fan is $N/b^k < b$ --- the same staircase discount
as \S7 --- and $M^*$ falls to ${\sim}0.2$--$3$\,KB (e.g.\ $W/D = 22$ at
$(N, D) = (10^6, 10^4)$, so $M^* \approx 0.4$\,KB). Theorem~2's headline
survives with its regime named: \textbf{hash traffic dominates for bodies
under ${\sim}5$--$13$\,KB when divergence is sparse ($D \lesssim b$), but
only under ${\sim}0.2$--$3$\,KB when it is heavy ($D \gg b$)}.

\paragraph{Numbers.} $W$ from \S7's table at $N = 10^6$ (so $C \approx N$),
$D_{\mathrm{red}} = 0$, $\bar{r} = 20$, $\ell = 33$, and a nominal
$s_v = 16$\,B; ``share'' is the hash channel's fraction of $E[B]$
(requests and framing are negligible at this scale):

\begin{center}
\begin{tabular}{@{}rrrrrrrr@{}}
\toprule
$C/D$ & $D$ & $W$ & hash bytes & $M^*$ &
\multicolumn{3}{c@{}}{hash share of $E[B]$} \\
\cmidrule(l){6-8}
& & & & & $M = 256$\,B & $M = 4$\,KB & $M = 64$\,KB \\
\midrule
${\sim}10^6$ & $1$    & $527$              & $10.5$\,KB & $10.5$\,KB & $97\%$ & $72\%$ & $14\%$ \\
${\sim}10^4$ & $10^2$ & $27\,\mathrm{k}$   & $0.54$\,MB & $5.4$\,KB  & $95\%$ & $57\%$ & $7.6\%$ \\
${\sim}10^2$ & $10^4$ & $220\,\mathrm{k}$  & $4.4$\,MB  & $0.39$\,KB & $59\%$ & $9.6\%$ & $0.7\%$ \\
$9$          & $10^5$ & $1.08\,\mathrm{M}$ & $22$\,MB   & $0.17$\,KB & $41\%$ & $5.0\%$ & $0.3\%$ \\
\bottomrule
\end{tabular}
\end{center}

Read against the document's running example: moving one differing message
through a $10^6$-leaf tree costs ${\approx}10.5$\,KB of hashes (\S4's
9\,KB hand count) however small the body --- half the pre-truncation
figure, which is exactly the Merkle-width halving banked; the residual
constant is the radix-256 tradeoff of \S10.6, not a fixable inefficiency.

\paragraph{Does the $D_L$/$D_R$ split matter?} For the totals, no: each
live differing leaf ships exactly once whichever side holds it, and $W$
already counts both directions' listings. The only split-sensitive
quantity in the derivation is Lemma~2's pair count
$D_L(C + D_R) + D_R(C + D_L) = DC + 2 D_L D_R$, which ranges from $DC$
(one-sided) to $DC + D^2/2$ (even split) --- at most a factor 2 inside the
$\log_b$, i.e.\ $\le \log_{256} 2 \approx 0.13$ levels of separation
depth. Per-direction loads do follow the split: the side holding $D_L$ of
the difference sends $D_L$ of the bodies, while the alternation splits the
hash records between the directions regardless.

\section{An experiment that distinguishes them}

All using existing infrastructure; no protocol changes.

\begin{enumerate}
\item \textbf{Byte counting (noise-free, decisive).} In
  \code{benches/gossip\_grid.rs}, wrap the
  \code{PipeReader}/\code{PipeWriter} ends of \code{Wire}
  (\code{gossip\_grid.rs:43-99}) in trivial
  \code{std::io::Read}/\code{Write} adapters that increment
  \code{AtomicU64}s, and record bytes per session per cell. Bandwidth is
  $\Theta(20 \cdot W + \text{body terms})$ by Theorem~2 and is exactly
  measurable. Regress $\log(\textit{bytes} - a \cdot D)$ on $\log D$ and
  $\log(N/D)$ over cells with $D \ge 100$: the sqrt model predicts
  elasticity $\approx 0.5$ in $\log(N/D)$; Theorem~1 predicts $\approx 0$
  (a log-of-log). Frame \emph{counts} from the same adapter (count 4-byte
  headers) verify Theorem~3 against
  $\tfrac{1}{2} \cdot \log_{256}(2DN) + c$ to $\pm 1$ frame.
\item \textbf{The flat decade.} Fix $\textit{differing} = 1{,}000$,
  $\textit{redacted} = 0$, sweep
  $\textit{common} \in \{10^3, 10^4, 10^5\}$ (cells already exist; extend
  \code{COMMON} in \code{benches/support/grid.rs:71-74} to $10^6$ for the
  decisive decade --- it is capped at $10^5$ today precisely because the
  per-iteration fixture build dominates, so accept a below-floor sample
  count or a one-off harness). Predicted wall-time
  ratios $\textit{common } 10^5 \to 10^6$: sqrt $\times 3.16$; Theorem~1
  $\times 1.2$ (the $0.08$-slope decade). One afternoon of Criterion time.
\item \textbf{Direct hash-compare counts.} Drive two
  \code{local::Exchange}s in-process through the \code{\#[cfg(test)]}
  \code{mirror::mirror} entry (\code{mirror.rs:384-400}) with grid-shaped
  trees and count \code{uncertain} entries (e.g.\ instrument in a test, or
  just count from the byte totals:
  $\textit{entries} \approx (\textit{uncertain bytes})/(18 \dots 48)$).
  This measures $W$ itself, bypassing the $\alpha \cdot D + \textit{fixed}$
  wall-time floor entirely.
\end{enumerate}

\section{Caveats and edge regimes}

\begin{enumerate}
\item \textbf{Lazy memoization (first session after writes).} A batch of
  $m$ local changes leaves cold \code{OnceLock}s on the
  $\Theta(m \cdot (1 + \log_b(N/m)))$ ancestor nodes of the changed paths
  (\code{untyped.rs:39-51}; memos survive \code{Arc}-sharing, so
  \emph{only} changed paths are cold, \code{untyped.rs:104-127}). The
  first \code{hash()} then pays one branch preimage per cold node --- up
  to $1 + 17 \cdot 256 \approx 4.4$\,KB of BLAKE3 for a saturated branch
  (\code{CHILD\_RECORD\_LEN}, \code{hash.rs:80-83}),
  but only 18 bytes per compressed-spine level (\code{untyped.rs:281-289}).
  \code{benches/hash\_recon.rs} measures both conventions at the
  \emph{pre-truncation} width (8\,KiB wraps vs.\ 34-byte commitments) and
  has not been re-cut to 17-byte child records, unlike
  \code{benches/branch\_hash.rs}; its comparison shape stands, its
  absolute numbers are ${\sim}2\times$ today's. This cost
  is additive, amortizes across sessions, and is excluded from the benches
  by \code{warm\_caches()} (\code{gossip\_grid.rs:154-160}).
\item \textbf{Bootstrap / one side empty ($D = N$).} No prefix has content
  on both sides, so there are \emph{zero} disputes: the empty responder
  sends an empty \code{Opening} (\code{local.rs:346-357}), and the
  initiator's root-level drain Left-provides every root child in one
  message (\code{partition.rs:282-294}), filtered through \code{Unknown}
  against the empty version (everything survives). Cost: $\Theta(N)$
  transfer and absorption, $O(1)$ rounds. Theorem~1's formula degrades
  gracefully to $\Theta(N)$ here, but the \emph{mechanism} is different
  --- the dispute machinery never engages.
\item \textbf{Deletion honoring.} A redacted leaf (present on A, deleted
  on B) keeps hashes differing along its path, so it costs descent exactly
  like a differing leaf --- \textbf{count redactions in $D$} (the grid's
  throughput accounting already does: \code{grid.rs:116-127}). But it
  costs no body bytes: at separation, A's Left-provide is killed by the
  $O(1)$ ceiling check (\code{unknown.rs:45-47}) and A deletes its own
  copy (\code{partition.rs:232-239}) --- deletion \emph{propagates}
  through the version ceiling, with no tombstone and no transfer. A whole
  missing-but-dominated subtree dies the same way: one memoized
  comparison, no descent below it.
\item \textbf{Skew.} A1 is enforced by content addressing, not hoped for:
  because the version is folded into the leaf path (\code{path.rs:32-50}),
  even a million-message single-party version-batch spreads uniformly ---
  concentration under few prefixes would require engineered BLAKE3
  outputs. (Contrast key-ordered Merkle trees, where a contiguous batch
  \emph{does} cluster.) Even granting an adversary clustered paths, the
  depth ceiling caps the damage at $W \le 32 \cdot b \cdot D$ and 18 RTT.
\item \textbf{Version sizes.} ITC \code{Version}s are not constant-size:
  they ride the handshake, every provided leaf, and every
  \code{ceiling/floor} $\le$ comparison (\code{unknown.rs:45, 60-63}),
  contributing the $s_v$ factors in Theorems~1--2. For long-lived
  universes with many party splits, $s_v$ is the term to watch, orthogonal
  to the tree combinatorics.
\item \textbf{The radix constant is the design tradeoff.} Per disputed
  node the protocol ships a full sibling fan (up to $256 \times 20$ish
  bytes $\approx$ 5\,KB --- the figure \code{mirror.rs}'s Cost section
  quotes) to buy a 2.4-decade depth reduction per level.
  $b = 256$ puts every realistic deployment within ${\sim}3$ levels of
  separation (RTT-optimal) at the cost of the $\times b$ in $W$ --- which
  is exactly the constant that made a $\sqrt{C \cdot D}$ fit land in the
  right numeric range in the first place.
\item \textbf{The two hash widths are a deliberate asymmetry.} Everything
  the protocol \emph{compares} is the 16-byte truncated Merkle hash;
  everything that \emph{identifies} is the full 32-byte content hash.
  The split is safe because the failure modes differ in kind: a Merkle
  false-equal (A7) prunes one divergent subtree only until the next
  mutation beneath that prefix perturbs every branch hash above it ---
  transient, self-healing, nothing dropped --- while a content-hash (path)
  collision would be permanent, silent split-brain. The path therefore
  keeps full width, and so does \emph{every component hash feeding it}: a
  16-byte inner $H(\textit{value})$ would cap the whole path at $2^{64}$
  despite its 32-byte output (\code{path.rs:32-50}). The truncation enters
  this analysis only through $\bar{r}$ and the branch-preimage constants
  --- it halved Theorem~2's hash term and \S8's crossovers, and changed
  no count. The full argument, adversarial cases included, lives in
  \code{hash.rs:24-58} and
  \code{results/2026-06-11-hash-width-transcript.md}.
\end{enumerate}

\end{document}
